Density-dependent selection and the limits of relative fitness

### 2017-12-22 manuscript.tex
l

Constant relative fitness values can be justified as a linear approximation \cite[pp. 277]{ewens_2004} \citep[Chap. 4]{charlesworth_1994} which is close to exact provided that selection is sufficiently weak and stable over time. Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}. The question is whether relative fitness can be used when selection is not vanishingly weak. In age-structured populations that reproduce by outcrossing, selection cannot in general be represented in terms of type-specific relative-fitness constants unless selection is weak \citep[Chap. 4]{charlesworth_1994}. We will therefore restrict our attention to asexual haploids with little or no age structure, where the relative fitness description may be valid even if selection is strong. This simplification also makes it easier to grasp the population ecological underpinnings of selection, as opposed to the role of mating systems or complex changes in age structure.

@@

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection. Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was necessary to assume that $s$ will be a few percent at most.

>>>>

Constant relative fitness values can be justified as a linear approximation \cite[pp. 277]{ewens_2004} \citep[Chap. 4]{charlesworth_1994} which is close to exact provided that selection is sufficiently weak and stable over time. Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}. The question is whether relative fitness can be used when selection is not vanishingly weak. In general, age-structured populations that reproduce by outcrossing do not permit strong selection to be represented in terms of type-specific relative-fitness constants \citep[Chap. 4]{charlesworth_1994}. We will therefore restrict our attention to asexual haploids with little or no age structure, where it is easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions.

@@

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection. Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most. While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_2014}.
